---
title: "Pseudobulk analysis of crypts"
author: "Ricardo Ramirez"
date: "2022-12-15"
output: 
  html_document:
      code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, message=FALSE}
library(tidyverse)
library(scuttle)
library(edgeR)
library(Seurat)
library(knitr)
library(rmarkdown)
library(ComplexHeatmap)
```

## Experimental design

We have been provided with groups of spots that belong to a certain crypt within a visium slide (n = 2). Within each slide there are two types of crypts (healthy and disease) and each slide represents a different condition.

The objective is to:

1) Identify the differentially expressed genes between types of crypts
2) Assess if there is an interaction with the different conditions (not detached from technical since there are no replicates of diets)

### Extract count matrix and meta data of original data

```{r}
d6_data <- readRDS("./R_objects/se.helminth.d6")
# Count data
DefaultAssay(d6_data) <- "RNA"
GEX_mat <- Seurat::GetAssayData(d6_data, assay = "RNA", slot ="counts")
GEX_mat <- GEX_mat[grepl("refdata-gex-mm10",rownames(GEX_mat)),]
rownames(GEX_mat) <- gsub("refdata-gex-mm10----------", "", rownames(GEX_mat))

# Meta data
meta_data <- d6_data@meta.data %>%
  rownames_to_column("raw_spot_id") %>%
  dplyr::mutate(spot_id = strsplit(raw_spot_id,
                                   "_") %>%
                  map_chr(., ~.x[[1]])) %>%
  dplyr::select(raw_spot_id, spot_id, diet, day, section_id)
```

### Spots of each crypt were provided

```{r, message=FALSE, warning= FALSE}
list_spots <- map(list.files("./ann/",full.names = T) %>% set_names,
    read_csv) %>%
  enframe() %>%
  unnest()

colnames(list_spots) <- c("file","spot_id", "ctrl_crypt", "granuloma_crypt")

list_spots <- list_spots %>% 
  dplyr::filter(!is.na(ctrl_crypt) | !is.na(granuloma_crypt)) %>%
  pivot_longer(-c(file,spot_id),names_to = "crypt_type", values_to = "id") %>%
  na.omit() %>%
  dplyr::mutate(diet = ifelse(grepl("STD", file), "STD", "GW")) %>%
  dplyr::select(-file)

meta_data <- meta_data %>%
  left_join(list_spots, by = c("spot_id", "diet")) %>%
  na.omit() %>%
  arrange(crypt_type, id) %>%
  dplyr::mutate(id = paste0(diet, "_", id))

# Check for duplicated spots during manual annotation
duplicated_spots <- meta_data$raw_spot_id[duplicated(meta_data$raw_spot_id)]

meta_data <- meta_data %>%
  dplyr::filter(!raw_spot_id %in% duplicated_spots)

coldata <- meta_data %>%
  dplyr::select(id, diet, crypt_type) %>%
  unique()
```

### Create pseudobulk profiles

```{r}
GEX_mat <- GEX_mat[,meta_data$raw_spot_id]

pb_GEX_mat <- scuttle::sumCountsAcrossCells(x = GEX_mat,
                                            ids  = meta_data %>% 
                                              dplyr::select(raw_spot_id, id) %>% 
                                              column_to_rownames("raw_spot_id") %>% 
                                              DataFrame())

pb_GEX_mat@colData <- pb_GEX_mat@colData %>%
  as.data.frame() %>%
  left_join(coldata, by = "id") %>%
  DataFrame()

assay(pb_GEX_mat, "counts") <- assay(pb_GEX_mat, "sum")
```

### Filter genes based on expression

Here we are strict in the sense that we have 6 pseudobulk profiles per diet and granuloma type, hence we will keep genes that have at least 1 read in 6 samples of a given group

```{r}
keep <- edgeR::filterByExpr(pb_GEX_mat, 
                            min.prop = 1, 
                            group = paste0(pb_GEX_mat@colData$diet,
                                           pb_GEX_mat@colData$crypt_type) %>%
                              as.factor(), 
                            min.count = 1, 
                            min.total.count = 6)

pb_GEX_mat <- pb_GEX_mat[keep,]
print("dimension of matrix genes/samples")
dim(pb_GEX_mat)
```

Do tmm normalization for regular exploratory analysis

```{r}
scale_factor = 100000
all_nf <- edgeR::calcNormFactors(pb_GEX_mat, method = "TMM")
sfs <- all_nf$samples$lib.size * all_nf$samples$norm.factors
pb <- sweep(assay(pb_GEX_mat, "counts"), MARGIN = 2, sfs, FUN = "/")
assay(pb_GEX_mat, "logcounts") <- log1p(pb * scale_factor)
```

# Using the results from DEA

```{r}
fit <- readRDS("./results/edgeR_fit.rds")
my_contrasts <- readRDS("./results/edgeR_contrasts.rds")
```

### Effect of granuloma crypt

What are the up or downregulated genes in the granuloma vs the ctrl crypt? This is done across slides

```{r}
granuloma_degs <- glmQLFTest(fit, contrast=my_contrasts[,"granulomavcrpt"])

granuloma_degs <- topTags(granuloma_degs, n = Inf) %>%
    as.data.frame() %>%
    rownames_to_column("gene") %>%
    dplyr::mutate(rank_stat = abs(logFC) * -log10(FDR))

g_degs_flexible <- granuloma_degs %>%
  dplyr::filter(FDR <= 0.1) %>%
  arrange(desc(rank_stat)) %>%
  pull(gene)
```

### Effect of diet in all groups

What are the up or downregulated genes in the GW vs the STD diet?

```{r}
diet_degs <- glmQLFTest(fit, contrast=my_contrasts[,"diet_effect"])

all_diet_degs <- topTags(diet_degs, n = Inf) %>%
    as.data.frame() %>%
    rownames_to_column("gene") %>%
  dplyr::filter(FDR <= 0.1)

diet_degs <- topTags(diet_degs, n = Inf) %>%
    as.data.frame() %>%
    rownames_to_column("gene") %>%
  dplyr::filter(FDR <= 0.1) %>%
  dplyr::mutate(rank_stat = abs(logFC) * -log10(FDR)) %>%
  arrange(desc(rank_stat)) %>%
  dplyr::slice(1:10) %>%
  pull(gene)
```

# Now effect of crypt in different diets

```{r}
std_diet_granuloma <- glmQLFTest(fit, contrast=my_contrasts[,"granuloma_effect_std"])

std_diet_granuloma <- topTags(std_diet_granuloma, n = Inf) %>%
    as.data.frame() %>%
    rownames_to_column("gene") %>%
    dplyr::filter(FDR <= 0.1) %>%
  dplyr::mutate(rank_stat = abs(logFC) * -log10(FDR)) %>%
  arrange(desc(rank_stat)) %>%
  dplyr::slice(1:10) %>%
  pull(gene)
```

```{r}
gw_diet_granuloma <- glmQLFTest(fit, contrast=my_contrasts[,"granuloma_effect_gw"])

gw_diet_granuloma <- topTags(gw_diet_granuloma, n = Inf) %>%
    as.data.frame() %>%
    rownames_to_column("gene") %>%
    dplyr::filter(FDR <= 0.1) %>%
  dplyr::mutate(rank_stat = abs(logFC) * -log10(FDR)) %>%
  arrange(desc(rank_stat)) %>%
  dplyr::slice(1:10) %>%
  pull(gene)
```

Final heatmap

```{r}
counts <- pb_GEX_mat %>% assay(., "logcounts") %>% t() %>% scale() %>% t()

ha <- HeatmapAnnotation(diet = colData(pb_GEX_mat)[,"diet"],
                        crypt_type = colData(pb_GEX_mat)[,"crypt_type"],
                        col = list(crypt_type = c("granuloma_crypt" = "yellow", "ctrl_crypt" = "pink"),
                                   diet = c("GW" = "darkgreen", "STD" = "orange")))


sign_genes <- unique(c(g_degs_flexible, diet_degs, gw_diet_granuloma, std_diet_granuloma))

pdf("./pub_results/hmap_edger_sign0.1.pdf", height = 6, width = 6)

ComplexHeatmap::Heatmap(counts[sign_genes,], name = "Expression", top_annotation = ha)

dev.off()
```

